\Large{Learning to Pivot with Adversarial Networks}

### 2016-10-31 tex/paper.tex
large font for fig2
larger figures
Reworked HEP example
I reworded the HEP example to have a little more motivation at the
beginning and a more developed connection to the hypothesis test,
motivation for AMS, motivation for pivoting on y=0, and optimization of
the hyper parameter lambda.

\begin{figure*}
    \begin{center}
        \includegraphics[width=0.23\textwidth]{figures/f-plain.pdf}
        \includegraphics[width=0.23\textwidth]{figures/surface-plain.pdf}
        \includegraphics[width=0.23\textwidth]{figures/f-adversary.pdf}
        \includegraphics[width=0.23\textwidth]{figures/surface-adversary.pdf}
    \end{center}
    \vspace{-0.5cm}
    \caption{Toy example.
    (Left) Conditional probability densities of the decision scores at $Z=-\sigma, 0, \sigma$
       when $f$ is built without adversarial training. The resulting densities
       are clearly dependent on the continuous parameter $Z$, indicating that $f$ is not pivotal.
    (Middle left) The corresponding decision surface, highlighting
       the fact that samples are easier to classify for values of $Z$ above $\sigma$,
       hence partially explaining the dependency.
    (Middle right) Conditional probability densities of the decision scores at $Z=-\sigma, 0, \sigma$ when $f$ is
       built with adversarial training, as outlined in Section~\ref{sec:method}.
       The resulting densities are now almost identical to each other, indicating only a
       small dependency on $Z$.
    (Right) The corresponding decision surface, illustrating how adversarial
       training bends the decision function vertically to erase the dependency on $Z$.
    }
    \label{fig:toy}
\end{figure*}

@@

Jets, or collimated sprays of particles produced by high energy quarks and
gluons, are produced ubiquitously at high energy colliders like the LHC \citep{LHCMachine}.
In a wide array of new physics models, new highly
massive particles can decay through known heavy particles in the Standard Model (SM),
like the $W$, $Z$, Higgs bosons, or top quarks which subsequently decay to
multiple quarks.  When these heavy SM particles have significant energy,
the Lorentz boost causes their quark decay products to merge into a single jet
which has a rich internal substructure (see e.g.
\citep{Altheimer:2012mn,Altheimer:2013yza} and references therein).
Distinguishing these boosted jets produced by heavy SM particles from
vanilla quarks and  gluons is a fundamental challenge in searching for signs of
new high mass particles. Challenging in its own right, this classification
problem is made all the more difficult by the presence of pileup, or multiple
simultaneous proton-proton interactions occurring at the same time as the primary
interaction.  These pileup interactions produce additional particles that can
contribute significant energies to jets unrelated to the underlying
discriminating information. Thus pileup acts as a source of noise in the
classification problem.

The classification challenge used here is common in jet substructure
studies (see \citep{Khachatryan:2014vla,ATL-PHYS-PUB-2015-033,wbosonATLAS} and
references therein): we aim to discriminate between jets produced by quark or gluon fragmentation (the
background) and  boosted $W$ boson jets containing a collimated pair of
quarks (the signal).  We reuse the datasets used in reference
\citep{baldi2016jet}.  Briefly summarizing, jets are built with the
anti-$k_T$ algorithm \citep{Cacciari:2008gp}
with radius parameter $R=1.2$, and trimmed \citep{Krohn:2009th} with subjets built with
the $k_T$ algorithm and parameter $f_{cut}=0.2$. The features used
in the classifiers are: trimmed jet invariant mass, N-subjettiness
$\tau_{21}^{\beta=1}$ \citep{nsub,Thaler2012}, and the energy correlation
functions \citep{Larkoski2013} $C_{2}^{\beta=1}$, $C_{2}^{\beta=2}$,
$D_{2}^{\beta=1}$, $D_{2}^{\beta=2}$.

As nuisance parameter, we consider events without pileup ($Z=0$) and
events with pileup ($Z=1$), for which an average number of $\langle \mu \rangle =
50$ unrelated additional pileup interactions are overlaid. Our goal is to build an accurate
classifier, for which we also want to minimize the effects due to the
uncertainties on the nuisance. More specifically, we choose to recast the
classification problem as a hypothesis test between signal+background (jets
originating from $W$ bosons or from quarks and gluons) and background
only (jets originating from quarks and gluons only) and tune the decision
threshold of our classifier by maximizing its approximate median significance
(AMS), when uncertainties in the background are taken into account (see Eqn.~20
of \citep{adam2014higgs}). Our motivation is that reducing the effects of
uncertainties by requiring independence of $Z$ with the classifier output $f(X;\theta_f)$ should
allow for a larger maximum significance.

To minimize the effects of $Z$ in the background events, we train a classifier
using  Algorithm~\ref{alg:adversarial-training} but consider the adversarial
term ${\cal L}_r$ conditioned on $Y=0$ only, as outlined in
Sec.~\ref{sec:problem}. Both $f$ and $r$ are neural networks with 3 hidden
layers of 64 nodes and ReLU activations, each terminated by a single final
output node with a sigmoid activation. Experiments are performed on a subset of
150000 samples for training while AMS is evaluated on an independent test set of
5000000 samples. Both training and testing samples are weighted such that the
signal-to-background ratio is 1/10, allowing us to probe the efficacy of the
method proposed here in a background dominated environment. Results reported
below are averages over 5 runs.

@@

As  Fig.~\ref{fig:physics-ams} illustrates, without adversarial training (at
$\lambda=0|Z=0$ when building a classifier at the nominal value $Z=0$ only, or
at $\lambda=0$ when building a classifier on data sampled from $p(X,Y,Z)$), the
maximum significance peaks at $7$. By contrast, as the independence constraint
is made stronger (for $\lambda > 0$) the AMS peak moves higher, with a maximum
value around  $7.8$ for $\lambda=10$. In other words, trading classification
accuracy for independence with respect to pileup results in a positive total net
benefit in terms of statistical significance. Setting $\lambda$ too high however
(e.g. $\lambda=500$) results in a decrease of the maximum significance, by
focusing the capacity of $f$ too strongly on independence, at the expense of
classification accuracy. As demonstrated in this example, controlling the
classification versus pivot trade-off through $\lambda$ therefore gives us a
principled and effective approach for maximizing significance by desensitizing
the classifier output $f(X;\theta_f)$ in the most beneficial way.

@@

To account for systematic uncertainties, experimentalists in high energy physics
typically take as fixed a classifier $f$ built from training data for a nominal
value $z_0$ of the nuisance parameter, and then propagate uncertainty
 by estimating $p(f(x)|z)$ with a parametrized calibration
procedure. Clearly, this classifier is however not optimal for $z \neq z_0$.
To overcome this issue, the classifier $f$ is sometimes built instead on a mixture
of training data generated from several plausible values $z_0, z_1, \dots$ of the nuisance parameter.
While this certainly improves classification performance with respect to the marginal model $p(X,Y)$,
there is no reason to expect the resulting classifier to be pivotal, as shown
previously in Section~\ref{sec:toy}.
As an alternative, parametrized
classifiers~\citep{cranmer2015approximating,Baldi:2016fzo} directly take
(nuisance) parameters as additional input variables, hence ultimately providing
the most statistically powerful approach for incorporating the effect of
systematics on the underlying classification task.  As argued in
\citep{Neal:2007zz}, it is not obvious how such classifiers can be used on real data since
the correct value $z$ of the nuisance often remains unknown. This is
not an issue in the context of parameter
inference~\citep{cranmer2015approximating}, where nuisance parameters are
eliminated via optimization or marginalization.
%, but otherwise often limits the range of their applications. 
In practice, parametrized classifiers  are  computationally expensive to build
and evaluate. In particular, calibrating their decision function, i.e.
approximating $p(f(x,z)|y,z)$ as a continuous function of $z$, remains an open
challenge. By contrast, constraining $f$ to be pivotal yields a classifier which
may not be optimal with respect to $Y|X,Z$, as discussed in
Section~\ref{sec:theory}, but that can otherwise be directly used in a wider range of
applications, since the dependence on the nuisance parameter $z$ has already been eliminated. 
Moreover, if $f$ is sufficiently close to being pivotal, then calibration only needs to be carried out once.

@@

\begin{acknowledgments}
    KC and GL are both supported through NSF ACI-1450310, additionally KC is supported
    through PHY-1505463 and PHY-1205376.
\end{acknowledgments}

>>>>

\begin{figure*}
    \begin{center}
        \includegraphics[width=0.245\textwidth]{figures/f-plain.pdf}
        \includegraphics[width=0.245\textwidth]{figures/surface-plain.pdf}
        \includegraphics[width=0.245\textwidth]{figures/f-adversary.pdf}
        \includegraphics[width=0.245\textwidth]{figures/surface-adversary.pdf}
    \end{center}
    \vspace{-0.5cm}
    \caption{Toy example.
    (Left) Conditional probability densities of the decision scores at $Z=-\sigma, 0, \sigma$
       when $f$ is built without adversarial training. The resulting densities
       are clearly dependent on the continuous parameter $Z$, indicating that $f$ is not pivotal.
    (Middle left) The corresponding decision surface, highlighting
       the fact that samples are easier to classify for values of $Z$ above $\sigma$,
       hence partially explaining the dependency.
    (Middle right) Conditional probability densities of the decision scores at $Z=-\sigma, 0, \sigma$ when $f$ is
       built with adversarial training, as outlined in Section~\ref{sec:method}.
       The resulting densities are now almost identical to each other, indicating only a
       small dependency on $Z$.
    (Right) The corresponding decision surface, illustrating how adversarial
       training bends the decision function vertically to erase the dependency on $Z$.
    }
    \label{fig:toy}
\end{figure*}

@@

%Experiments at high energy colliders like the LHC \citep{LHCMachine}
%are searching for evidence of new particles beyond the ones
%described by the so-called Standard Model (SM) of particle physics.
%A wide array of extended theories predict the existence of new massive
%particles that would decay to known heavy particles in the Standard Model,
%like the $W$, $Z$, Higgs boson, or top quark, which subsequently decay to
%multiple quarks.
%Finally, those quarks would produce collimated sprays of particles known as \textit{jets}.
%However, jets are produced ubiquitously at high energy colliders through
%more mundane processes in the SM, which leads to a challenging classification problem
%that is beset with a number of sources of systematic uncertainty.

Experiments at high energy colliders like the LHC \citep{LHCMachine}
are searching for evidence of new particles beyond the ones
described by the well established Standard Model (SM) of particle physics.
%Searches for these new particles in the LHC data involve a challenging
%classification problem that is beset with a number of sources of systematic uncertainty.
%
A wide array of extended theories predict the existence of new massive
particles that would decay to known particles in the SM such as the $W$ boson.
The $W$ boson is unstable and can decay to
two quarks, each of which produce collimated sprays of particles known as \textit{jets}
(collectively denoted $W\to jj$).
If the exotic particle is heavy, then the $W$ boson will be moving very fast,
and  relativistic effects will cause the two jets from its decay to merge into a single `fat jet'.
These fat jets have a rich internal substructure (see e.g.
\citep{Altheimer:2012mn,Altheimer:2013yza} and references therein).
However, jets are produced ubiquitously at high energy colliders through
more mundane processes in the SM, which leads to a challenging classification problem
that is beset with a number of sources of systematic uncertainty.
%Distinguishing these ($y=1$) fat jets produced by heavy SM particles from
%the copious background of normal ($y=0$) jets is a fundamental challenge in searching
%for signs of new high mass particles.

%A wide array of extended theories predict the existence of new massive
%particles that would decay to known heavy particles in the Standard Model
%(e.g. the $W$ and $Z$ bosons), which subsequently decay to
%multiple quarks.
%Those quarks subsequently produce collimated sprays of particles known as \textit{jets} (denoted $j$).
%However, jets are produced ubiquitously at high energy colliders through
%more mundane processes in the SM, which leads to a challenging classification problem
%that is beset with a number of sources of systematic uncertainty.

%Consider the binary classification problem where $y=0$ corresponds to data produced from the
%ubiquitous SM process $pp \to jj$ and $y=1$ corresponds to data produced from an exotic
%process $pp \to W' \to W(\to jj) Z(\to jj)$ predicted by an extended theory.
%When the mass of the $W'$ is much larger than the mass of the $W$ and $Z$ bosons,
%then the $W$ and $Z$ particles are moving very fast and the relativistic Lorentz boost
%causes the two jets from their decay to merge into a single `fat jet',

%Consider the search for an exotic particle whose decay leads to a $W$ boson
%that subsequently decays to two jets via $W\to jj$.  If the $W$ boson is moving
%very fast, then the relativistic Lorentz boost
%causes the two jets from its decay to merge into a single `fat jet',
%which has a rich internal substructure (see e.g.
%\citep{Altheimer:2012mn,Altheimer:2013yza} and references therein).
%%We can cast this as a binary classification problem where $y=0$
%%corresponds to data produced from the ubiquitous SM process
%%that produce jets and $y=1$  corresponds to data produced from an exotic
%%process that gives rise to fat jets from $W\to jj$.

%binary classification problem where $y=0$ corresponds to data produced from the
%ubiquitous SM process $pp \to jj$ and $y=1$ corresponds to data produced from an exotic
%process $pp \to W' \to W(\to jj) Z(\to jj)$ predicted by an extended theory.
%When the mass of the $W'$ is much larger than the mass of the $W$ and $Z$ bosons,
%then the $W$ and $Z$ particles are moving very fast and the relativistic Lorentz boost
%causes the two jets from their decay to merge into a single `fat jet',
%which has a rich internal substructure (see e.g.
%\citep{Altheimer:2012mn,Altheimer:2013yza} and references therein).

%Distinguishing these ($y=1$) fat jets produced by heavy SM particles from
%the copious background of normal ($y=0$) jets is a fundamental challenge in searching
%for signs of new high mass particles.

The classification challenge used here is common in jet substructure
studies (see \citep{Khachatryan:2014vla,ATL-PHYS-PUB-2015-033,wbosonATLAS} and
references therein): we aim to discriminate between  the normal jets produced copiously
at the LHC ($y=0$) and  the fat jets coming from an exotic process ($y=1$).
We reuse the datasets used in reference
\citep{baldi2016jet}.  For completeness, the jets are built with the
anti-$k_T$ algorithm \citep{Cacciari:2008gp}
with radius parameter $R=1.2$, and trimmed \citep{Krohn:2009th} with subjets built with
the $k_T$ algorithm and parameter $f_{cut}=0.2$. The features used
in the classifiers are: trimmed jet invariant mass, N-subjettiness
$\tau_{21}^{\beta=1}$ \citep{nsub,Thaler2012}, and the energy correlation
functions \citep{Larkoski2013} $C_{2}^{\beta=1}$, $C_{2}^{\beta=2}$,
$D_{2}^{\beta=1}$, $D_{2}^{\beta=2}$.

Challenging in its own right, this classification
problem is made all the more difficult by the presence of pileup, or multiple
 proton-proton interactions occurring simultaneously with the primary
interaction.  These pileup interactions produce additional particles that can
contribute significant energies to jets unrelated to the underlying
discriminating information. The number of pileup interactions can vary with
the running conditions of the collider, and we want the classifier to be robust
to these conditions. Taking some liberty, we consider an extreme case with
a categorical nuisance parameter, where $Z=0$ corresponds to events without pileup
and $Z=1$ corresponds to events with pileup, for which there are an average  of
50 independent pileup interactions overlaid.

We do not expect that we will be able to find a function $f$ that simultaneously
minimizes the classification loss $L_f$ and is pivotal. Thus, we need to optimize
the hyperparameter $\lambda$ of Eqtn.~\ref{eqn:vf-lambda} with respect to
a higher-level objective. In this case, the natural higher-level context is a
hypothesis test of a null hypothesis with no $Y=1$ events against an
alternate hypothesis that is a mixture of $Y=0$ and $Y=1$ events.
In the absence of systematic uncertainties optimizing $L_f$ simultaneously
optimizes the power of a classical hypothesis test in the Neyman-Pearson sense.
When we include systematic uncertainties we will need to balance the
classification performance against the robustness to uncertainty in $Z$.
Since we are still performing a hypothesis test against the null, we only
wish to impose the pivotal property on $Y=0$ events. Furthermore,
we use the Approximate Median Significance (AMS), that was used
in the Higgs boson machine learning challenge that is a natural generalization
of the power of a hypothesis test when systematic uncertainties are taken into account
(see Eqn.~20 of \citep{adam2014higgs}).

%Our goal is to build an accurate classifier, for which we also want to minimize the effects due to the
%uncertainties on the nuisance. More specifically, we choose to recast the
%classification problem as a hypothesis test between signal+background (jets
%originating from $W$ bosons or from quarks and gluons) and background
%only (jets originating from quarks and gluons only) and tune the decision
%threshold of our classifier by maximizing its approximate median significance
%(AMS), when uncertainties in the background are taken into account (see Eqn.~20
%of \citep{adam2014higgs}). Our motivation is that reducing the effects of
%uncertainties by requiring independence of $Z$ with the classifier output $f(X;\theta_f)$ should
%allow for a larger maximum significance.

For several values of $\lambda$, we train a classifier
using  Algorithm~\ref{alg:adversarial-training} but consider the adversarial
term ${\cal L}_r$ conditioned on $Y=0$ only, as outlined in
Sec.~\ref{sec:problem}. Both $f$ and $r$ are neural networks with 3 hidden
layers of 64 nodes and ReLU activations, each terminated by a single final
output node with a sigmoid activation. Experiments are performed on a subset of
150000 samples for training while AMS is evaluated on an independent test set of
5000000 samples. Both training and testing samples are weighted such that the
signal-to-background ratio is 1/10, allowing us to probe the efficacy of the
method proposed here in a background dominated environment. Results reported
below are averages over 5 runs.

@@

As  Fig.~\ref{fig:physics-ams} illustrates, without adversarial training (at
$\lambda=0|Z=0$ when building a classifier at the nominal value $Z=0$ only, or
at $\lambda=0$ when building a classifier on data sampled from $p(X,Y,Z)$), the
AMS peaks at $7$. By contrast, as the pivotal constraint
is made stronger (for $\lambda > 0$) the AMS peak moves higher, with a maximum
value around  $7.8$ for $\lambda=10$. In other words, trading classification
accuracy for robustness to pileup results in a net
benefit in terms of the power of the resulting hypothesis test. Setting $\lambda$ too high however
(e.g. $\lambda=500$) results in a decrease of the maximum AMS, by
focusing the capacity of $f$ too strongly on independence with $Z$, at the expense of
classification accuracy. As demonstrated in this example, optimizing $\lambda$
gives us a principled and effective approach to control the tradeoff between
classification accuracy and robustness that ultimately maximizes the
power of the enveloping hypothesis test.
%significance by desensitizing
%the classifier output $f(X;\theta_f)$ in the most beneficial way.

@@

To account for systematic uncertainties, experimentalists in high energy physics
typically take as fixed a classifier $f$ built from training data for a nominal
value $z_0$ of the nuisance parameter, and then propagate uncertainty
 by estimating $p(f(x)|z)$ with a parametrized calibration
procedure. Clearly, this classifier is however not optimal for $z \neq z_0$.
To overcome this issue, the classifier $f$ is sometimes built instead on a mixture
of training data generated from several plausible values $z_0, z_1, \dots$ of the nuisance parameter.
While this certainly improves classification performance with respect to the marginal model $p(X,Y)$,
there is no reason to expect the resulting classifier to be pivotal, as shown
previously in Section~\ref{sec:toy}.
As an alternative, parametrized
classifiers~\citep{cranmer2015approximating,Baldi:2016fzo} directly take
(nuisance) parameters as additional input variables, hence ultimately providing
the most statistically powerful approach for incorporating the effect of
systematics on the underlying classification task.  As argued in
\citep{Neal:2007zz}, it is not obvious how such classifiers can be used on real data since
the correct value $z$ of the nuisance often remains unknown. This is
not an issue in the context of parameter
inference~\citep{cranmer2015approximating}, where nuisance parameters are
eliminated via optimization or marginalization.
%, but otherwise often limits the range of their applications.
In practice, parametrized classifiers  are  computationally expensive to build
and evaluate. In particular, calibrating their decision function, i.e.
approximating $p(f(x,z)|y,z)$ as a continuous function of $z$, remains an open
challenge. By contrast, constraining $f$ to be pivotal yields a classifier which
may not be optimal with respect to $Y|X,Z$, as discussed in
Section~\ref{sec:theory}, but that can otherwise be directly used in a wider range of
applications, since the dependence on the nuisance parameter $Z$ has already been eliminated.
Moreover, if $f$ is sufficiently close to being pivotal, then calibration only needs to be carried out once.

@@

\begin{acknowledgments}

    KC and GL are both supported through NSF ACI-1450310, additionally KC is supported
    through PHY-1505463 and PHY-1205376.

\end{acknowledgments}

\vfill
